Appendix S1. Linear regression models and non-linear population dynamics

### 2024-05-04 Appendix_S1_v01.Rmd
began LV competition, observed bifurcation

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.67
r2.0 <- 1.7
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125. Relationship between E and Nt is also shown."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
```
## Linear statistical model with environmental covariate
XX
## Bayesian linear statistical model: HMSC
XX
## Conclusions
XX

>>>>

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.67
r2.0 <- 1.7
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125. Relationship between E and Nt is also shown."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
```
## Linear statistical model with environmental covariate

We fit each species' population time series to an arimaX and lm with population size of the others species as a covariate.

$$ N_{1,t} = \beta_0 + \beta_1 N_{1,t-1} + \beta_2 N_{2,t-1}+ \epsilon_t $$

```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125. Relationship between E and Nt is also shown."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N$N1,N$N2))
colnames(dat) <- c("N1","N2")
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N1)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, N2)) + geom_point()
p1 + p2
```

```{r,warning=FALSE}
## Fit the model for Species 1
m.3.ar.n1 <- arima(x = log(N$N1), order = c(1,0,0), include.mean = T, method="CSS", xreg=N$N2)
m.3.lm.n1 <- lm(log(dat$N1[2:t]) ~ log(dat$N1[1:(t-1)]) + log(dat$N2[1:(t-1)]))
#plotting the series along with the fitted values
m.3.ar.fit.n1 <- log(N$N1) - residuals(m.3.ar.n1)
m.3.lm.fit.n1 <- log(dat$N1[2:t]) - m.3.lm.n1$resid
dat$ar3.fit.n1 <- m.3.ar.fit.n1
dat$lm3.fit.n1 <- NA
dat$lm3.fit.n1[2:t] <- m.3.lm.fit.n1
## Species 2
m.3.ar.n2 <- arima(x = log(N$N2), order = c(1,0,0), include.mean = T, method="CSS", xreg=N$N1)
m.3.lm.n2 <- lm(log(dat$N2[2:t]) ~ log(dat$N2[1:(t-1)]) + log(dat$N1[1:(t-1)]))
#plotting the series along with the fitted values
m.3.ar.fit.n2 <- log(N$N2) - residuals(m.3.ar.n2)
m.3.lm.fit.n2 <- log(dat$N2[2:t]) - m.3.lm.n2$resid
dat$ar3.fit.n2 <- m.3.ar.fit.n2
dat$lm3.fit.n2 <- NA
dat$lm3.fit.n2[2:t] <- m.3.lm.fit.n2
```

```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
p1 <- ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N1),color="data")) +
  geom_point(aes(y=ar3.fit.n1, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm3.fit.n1, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N1), color="data"), linewidth=.4) +
  geom_line(aes(y=ar3.fit.n1, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm3.fit.n1, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N1") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
p2 <- ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N2),color="data")) +
  geom_point(aes(y=ar3.fit.n2, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm3.fit.n2, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N2), color="data"), linewidth=.4) +
  geom_line(aes(y=ar3.fit.n2, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm3.fit.n2, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N2") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
p1+p2
```

```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
dN1 <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN1[i] <- log(N$N1[i+1])-log(N$N1[i])
}
d.dN1 <- as.data.frame(dN1)
d.dN1$Nt <- log(dat$N1[1:(t-1)])
d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
p1 <- ggplot2::ggplot(d.dN1, aes(x=Nt,y=Nt_plus_1)) + geom_point()

d.dN1 <- as.data.frame(dN1)
d.dN1$Nt <- log(dat$N1[1:(t-1)])
d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
d.dN1$N2 <- log(dat$N2[2:t])
p2 <- ggplot2::ggplot(d.dN1, aes(x=N2,y=Nt_plus_1)) + geom_point()
p3 <- ggplot2::ggplot(d.dN1, aes(Nt, dN1)) + geom_point()
p4 <- ggplot2::ggplot(d.dN1, aes(N2, dN1)) + geom_point()

XXXXX

```{r}
df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
colnames(df) <- c("Nt1","N1","N2")
m.3.lm <- lm(Nt1 ~ N1 + N2,data=df)
```

```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
ggplot(data = df, aes(x = N2, y = Nt1)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x)
```
UH OH !!!!

## Bayesian linear statistical model: HMSC
XX
## Conclusions
XX
